On the Minimum Core Mass for Giant Planet Formation through Kelvin-Helmholtz Contraction

### 2013-05-27 tex/draft/draft2.tex
updated draft

\section{Introduction \textbf{(Paper II)}}
\label{intro}

@@

\section{Atmosphere Models \textbf{(Paper I)}}
\label{sec2}

In this section we derive the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk in the absence of planetesimal accretion. In section \ref{model} we describe the assumptions of our model. In section \ref{scales} we describe the properties of our assumed protoplanetary disk and review relevant length scales. Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere and the boundary conditions that we use. In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools. Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.

@@

\subsection{Disk Model and Length Scales}
\label{scales}

As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

We further define several characteristic length scales that are important for this problem. \\

\begin{equation}
\label{eq:H}
H \equiv \frac{c_s^2}{\Omega},
\end{equation}

@@

\subsubsection{Boundary Conditions}
\label{BCs}

@@

As stated in section \ref{model}, the atmosphere matches smoothly onto the disk at the outer boundary. The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$.

@@

\subsection{Quasi-Static Two-Layer Model}
\label{twolayer}

As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf. \ref{model}).

We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$. The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$.

\section{Analytic Cooling Model \textbf{(Paper I)}}
\label{analytic}

@@

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4}.
\end{equation}

@@

\section{Quasi-Static Kelvin-Helmholtz Contraction \textbf{(Paper I)}}
\label{KH}

In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models. As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$. We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU. As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$. We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$. We further explore how the atmosphere evolution varies with adiabatic index and molecular weight. We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$). The comparison between different gas compositions is deferred to section \ref{critcore}.

Figure 1 shows instantaneous atmosphere radial profiles at $a=10$ AU accreting around a fixed core of mass $M_{\rm{c}}=5 M_{\oplus}$ for several total planet masses. In what follows,  if the $R_B<R_H$ hierarchy holds, we define the planet mass as the total mass enclosed within the Bondi radius of the atmosphere, and regard the gas between the Bondi and Hill radii as perturbations due to the gravitational influence of the planet. For illustrative purposes we mark the position of the Bondi radius and of the radiative convective boundary. We notice that the pressure and temperature in the convective regions of the atmospheres scale as $1/r$, which is the expected behavior of an ideal gas adiabat with a polytopic equation of state (e.g., \citealt{rafikov06}). In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of the analytic model (see section \ref{analytic} and \App{sec:analytic}). In addition, the atmosphere temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, and so the atmosphere is nearly isothermal throughout the radiative zone. This is also in agreement with the analytic predictions. %, and justifies

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Example pressure, temperature and mass profiles as a function of radius.}
\end{figure}

The location of the Bondi radius moves further out as the mass of the atmosphere increases, as the Bondi radius scales with mass. The location of the radiative-convective boundary has a more interesting behavior. As stated in section \ref{intro}, the nebular gas initially settles into equilibrium around the protoplanetary core on a short timescale relative to the thermal timescale, and is isentropic throughout with an entropy equal to that of the disk. In the context of our model, this atmosphere represents the minimum atmosphere mass for which a hydrostatic solution exists. This atmosphere is purely convective. As gas accumulates and the atmosphere starts cooling, an outer radiative region forms. This layer is thin for low mass atmospheres, as can be seen in Fig. 1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low. An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract. As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e. the radiative regions of these types of atmospheres are getting deeper and deeper.  \textbf{[This needs to be rephrased, somewhat vague and repetitive.]}

We further use the global cooling model developed in section \ref{cooling} to obtain an evolutionary series from static atmosphere profiles. Figure 2 shows the luminosity and cooling time evolution with atmosphere mass. Gas accretion is initially rapid, but slows down significantly after a relatively small mass increase. For comparison, we also plot the analytic results derived in section \ref{analytic}. We see that there is only a very brief agreement between the analytic and full numerical results. At early times, the radiative zone is too shallow and so the approximation that $P_{\rm RCB} \gg P_{\rm d}$ used in section \ref{analytic} breaks down. At late times, the self-gravity of the atmosphere becomes important. %We see that self-gravity starts to matter

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution with mass. The analytic result is plotted for comparison.}
\label{fig:Ltplot}
\end{figure}

@@

The right-hand side of equation (\ref{eq:dti}) decreases with increasing atmosphere mass, and eventually becomes negative when the mass of the envelope becomes comparable to the core mass $M_{\rm{atm}} \sim M_{\rm{c}}$. This implies a negative luminosity from the RCB. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the RCB. The heat flux due to convective motions is always directed outwards, and hence it is positive. Moreover, the temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside. As a result (see equation (\ref{eq:structd})), the heat flux due to radiative diffusion is also positive. This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary.

@@

\section{Critical Core Mass \textbf{(Paper I, except for $\delad=2/5$ case)}}
\label{critical}

@@

Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere. The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences. Therefore, for a set of disk and gas conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass. %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core. This is the maximum mass the core can have before unstable gas accretion commences.

In our model the atmosphere is no longer in steady state, because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale. Figure \ref{fig:growthtime} shows the evolution of the atmosphere growth time with mass at several locations in our fiducial disk for a fixed core mass. In the early stages when $M_{\rm{atm}}$ is small, the envelope has a thin radiative zone and hence a high luminosity (see Figure \ref{fig:Ltplot}). As a result, the envelope initially grows quickly, as shown in Figure \ref{fig:growthtime}. As the radiative zone becomes thicker, the luminosity decreases and growth slows down. For our choice of parameters, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$. This is when the self-gravity of the atmosphere starts to matter; the envelope becomes massive enough for its energy to be large compared to the internal energy of the nebular gas. As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass. Since growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$ for the cases that we tried, the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude. We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices. We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass as defined above, and runaway accretion commences.

In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk. In section \ref{Mct} we show the dependence of the crossover time on the core mass at a fixed distance in the disk, justifying the need for a ``minimum'' core mass. We further explore how the crossover time is affected by the position in the disk in section \ref{TPeffects} and by the mean molecular weight and opacity of the nebular gas in section \ref{muopacity}. Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.

@@

\subsection{Influence of Core Mass on Crossover Time}
\label{Mct}

We begin by fixing the disk properties and generating atmosphere profiles for a set of fixed core masses. Figure \ref{fig:tvsM} shows the crossover time behavior for a range of cores with masses between 1 and 10 $M_{\oplus}$, at $a=10$ AU in our fiducial disk. The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity. This can also be seen from the analytical results in equation (\ref{eq:tcoolan}). As such, for a fixed crossover time, e.g. the typical lifetime of the protoplanetary disk, there is a minimum core mass needed to form a planet within this time frame, which we defined above as the critical core mass. We defer numerical estimates of the critical core mass to section \ref{critcore}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
%\vspace{-0.5in}
\caption{We generate atmosphere profiles at $a=10$ AU for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$. We plot the cumulative cooling time as a function of atmosphere mass. The circles mark the crossover time. A larger core mass results in a shorter crossover time.}
\label{fig:tvsM}
\end{figure}

\subsection{Role of Disk Temperature and Pressure}
\label{TPeffects}

In what follows we explore how the crossover time is affected by the planet's location in the disk. The location only matters because it dictates the disk temperature (equation (\ref{eq:diskb})) and pressure (equation (\ref{eq:Pd})). We therefore study the separate effects of disk temperature and pressure on the time evolution of the atmosphere. At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we fix the disk temperature $T_{\rm{d}}$ at the value given by equation (\ref{eq:diskb}) and vary the disk pressure $P_{\rm{d}}$. We perform a similar analysis on the effect of disk temperature $T_{\rm{d}}$ on the crossover time by fixing the disk pressure at the value given by equation (\ref{eq:Pd}) and varying $T_{\rm{d}}$. The resulting time evolution is plotted in Figure \ref{fig:TPeffects}. For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}. However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long. Here we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model. We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$.

\begin{figure*}[tb]
\centering
\includegraphics[width=1.\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/TdPd_effect.pdf}
%\vspace{-0.5in}
\caption{Left panel: crossover time as a function of disk temperature for fixed disk pressure and a core of fixed mass $M_{\rm c}=5 M_{\oplus}$. Right panel: crossover time as a function of disk pressure for fixed disk temperature and a core of fixed mass  $M_{\rm c}=5 M_{\oplus}$. The analytic scalings are plotted for comparison, for an effective crossover mass $M_{\rm{co}}=f M_{\rm c}$ with $f=0.13$. Gas accretion is faster for lower temperatures and higher pressures.} 
\label{fig:TPeffects}
\end{figure*}

>>>>

\section{Introduction}
\label{intro}

@@

\section{Atmosphere Models}
\label{sec2}

In this section we derive the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk in the absence of planetesimal accretion. In section \ref{model} we describe the assumptions of our model. Section \ref{struct} presents the equations describing the structure of a static planetary atmosphere.  In section \ref{BCs} we describe discuss our boundary conditions and the properties of our assumed protoplanetary disk, and review relevant length scales.  and the boundary conditions that we use. In section \ref{cooling} we develop a model to estimate the rate at which the atmosphere cools. Finally, we combine these calculations into a quasi-static model for atmospheric evolution in section \ref{twolayer}.

@@

%\subsection{Length Scales}
%\label{scales}

@@

\subsection{Boundary Conditions}
\label{BCs}

We first define several characteristic length scales that are important for this problem. \\

\begin{equation}
\label{eq:H}
H \equiv \frac{c_s}{\Omega},
\end{equation}

@@

As stated in section \ref{model}, the atmosphere matches smoothly onto the disk at the outer boundary. The temperature and pressure at the Hill radius have to therefore be given by the nebular temperature and pressure: $T(R_H)=T_d$ and $P(R_H)=P_d$. As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

@@

\section{Analytic Cooling Model}
\label{analytic}

@@

\begin{equation}
\label{eq:delo}
\nabla \di \equiv \frac{3 \kappa_0 P_{\rm d}}{64 \pi G M \sigma T_{\rm d}^4} L.
\end{equation}

@@

\section{Quasi-Static Kelvin-Helmholtz Contraction}
\label{KH}

In this section we generate quasi-static two-layer atmosphere profiles by integrating the equations of hydrostatic balance (\ref{eq:struct}) with the boundary conditions described in section \ref{BCs}. We obtain an evolutionary sequence from a series of instantaneous atmosphere profiles using the cooling model outlined in section \ref{cooling}. We describe our numerical method in section \ref{twolayer}. In section \ref{profiles} we show example plots of instantaneous and evolutionary atmosphere profiles for various choices of disk and core parameters. We compare our results with the predictions of the analytic model described in section \ref{analytic} and we explore the effect of changes in opacity on the atmosphere evolution. Finally, we discuss the breakdown of the quasi-static model in section \ref{endoftime}.

\subsection{Numerical Method}
\label{twolayer}

As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. In this section we describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf. \ref{model}).

We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are compared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_{\rm H}$, with the boundary conditions $T(R_{\mathrm{out}})=T_{\rm d}$ and $P(R_{\mathrm{out}})=P_{\rm d}$. The numerical integration gives a value for the core mass implied by the trial solution $M_{\rm c}=M_{\rm{c, guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{\rm{c,guess}}$ converges to the actual core mass $M_{\rm c}$.

\subsection{Atmosphere Profiles}
\label{profiles}

We consider planets embedded in a protoplanetary disk around a Sun-like star and with a temperature and pressure profile described by equations (\ref{eq:diskb}) and (\ref{eq:Pd}), respectively. We are primarily interested in the outer regions of the disk, so we generate atmosphere models at distances between 10 and 100 AU in our fiducial disk. We emphasize that the location in the disk only affects the temperature and pressure of the nebular gas. As our model applies only when planetesimal accretion occurs too slowly to substantially change the mass of the core, we consider the evolution of an atmosphere around a core of fixed mass.

We assume that the atmosphere consists of an ideal gas described by a polytropic equation of state, and is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$, resulting in a mean molecular weight $\mu=2.35$. We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic gradient of a diatomic gas $\delad=2/7$.  %We further explore how the atmosphere evolution varies with adiabatic index and molecular weight. We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$). The comparison between different gas compositions is deferred to section \ref{critcore}.

Figure \ref{fig:profiles} shows instantaneous radial profiles at different stages in the evolution of an atmosphere accreting around a core of fixed mass $M_{\rm{c}}=5 M_{\oplus}$ located at $a=60$ AU.  We define the atmosphere mass as the mass of the gas enclosed within the Bondi radius. The locations of the radiative-convective boundary and of the Bondi radius are displayed. Pressure and temperature in the convective region of the atmosphere scale as $1/r$, which is the expected behavior of an ideal gas adiabat (e.g., \citealt{rafikov06}). In the outer radiative region of the atmosphere, the pressure has a nearly exponential profile, which is in agreement with the predictions of our analytic model (see section \ref{analytic} and \App{sec:analytic}). In addition, the temperature at the radiative-convective boundary $T_{\rm{RCB}}$ only differs from the disk temperature $T_{\rm{d}}$ by an order unity factor, hence the atmosphere is nearly isothermal throughout the radiative zone. This is also in agreement with our analytic predictions (see equation \ref{eq:Tcb}). %, and justifies

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/PTm_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Example pressure, temperature and total mass (core plus atmosphere) profiles as a function of radius at $a=60$ AU and for a fixed core mass $M_{\rm c}=5 M_{\oplus}$. The circles and triangles mark the locations of the Bondi radius and of the radiative-convective boundary, respectively. The solid, dotted and dashed lines correspond to profiles with total mass (core and atmosphere) of $5.12 M_{\oplus}$, $5.85 M_{\oplus}$ and $9.22 M_{\oplus}$, respectively.}
\label{fig:profiles}
\end{figure}

In the initial stages of its evolution, the atmosphere is purely convective and isentropic, with an entropy equal to that of the disk. In the context of our numerical model, this is the minimum atmosphere mass for which a hydrostatic solution exists. Further gas accumulation brings in energy that has to be radiated away, and an initially thin outer radiative region forms, which becomes thicker as the atmosphere becomes more massive.

%This layer is thin for low mass atmospheres, An increase in atmosphere mass causes the radiative zone to expand and the convective region to contract. As more and more gas is accumulated, the convective zone eventually starts expanding again; however, the ratio between the radii of the radiative and convective regions continually increases with atmosphere mass, i.e. the radiative regions of these types of atmospheres are getting deeper and deeper.  \textbf{[This needs to be rephrased, somewhat vague and repetitive.]}

%as can be seen in Fig. 1: for the $5.21 M_{\oplus}$ atmosphere, the ratio between the thickness of the radiative and convective regions is low.

In what follows we discuss the time evolution of the envelope. We use the global cooling model developed in section \ref{cooling} to obtain an evolutionary history from series of static atmosphere profiles, using the numerical procedure described in section \ref{twolayer}. Figure \ref{fig:Ltplot} shows the luminosity evolution and the elapsed time as a function of atmosphere mass for the same choice of parameters as in Figure \ref{fig:profiles} ($M_{\rm c}=5 M_{\oplus}$). The radiative region of the atmosphere is initially thin, which results in a relatively small number of steps that photons need to diffuse and hence a high luminosity (see equation \ref{eq:Lcb}). The luminosity decreases as the atmosphere mass grows and the radiative region becomes deeper. For comparison, we also plot the analytic results derived in section \ref{analytic}. We find only a brief agreement between the analytic and numerical results. In the early stages, the radiative zone is too shallow and the assumption that $P_{\rm RCB} \gg P_{\rm d}$ used in the analytic model breaks down. As the mass of the atmosphere grows, the neglect of self-gravity in the analytic model results in a lower luminosity (equation \ref{eq:Lcb}) and therefore a longer cooling time when compared to the numerical solution. While our analytic model does not provide an accurate quantitative description of the atmosphere evolution, we will show in section \ref{critical} that it predicts the correct scalings for the atmosphere crossover time and critical core mass.  %We see that self-gravity starts to matter

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/Lt_profiles_v2.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution of the atmosphere of a planet with fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=60$ AU. The analytic result is plotted for comparison. The luminosity is initially high, then decreases as the atmosphere grows in mass and the radiative zone becomes deeper. Growth is accelerated as a result.}
\label{fig:Ltplot}
\end{figure}

We explain the time evolution from the bottom panel of Figure \ref{fig:Ltplot} by showing how the atmosphere growth time evolves with the atmosphere mass for a fixed core. Since the luminosity is high in the early stages, the envelope initially grows quickly; this is shown in Figure \ref{fig:growthtime} for a planet forming around a core of mass $M_{\rm c}=5 M_{\oplus}$ at several locations in our fiducial disk. Even though the atmosphere growth time, i.e. $M/\dot{M}$, is small in the early evolution stages, the increase in atmosphere mass, i.e. $\dot{M}$ is also small. As a result, the envelope spends a relatively long time at a low mass, which explains the behavior in the bottom panel of Figure \ref{fig:Ltplot}. As the radiative zone becomes thicker, the luminosity decreases and growth slows down. For $M_{\rm c}=5 M_{\oplus}$, we see that accretion is slowest for an atmosphere mass of $1-2 M_{\oplus}$. This is when the self-gravity of the atmosphere starts to matter; the envelope becomes massive enough for its energy to be large compared to the internal energy of the nebular gas. As a result, growth is continuously accelerated, and eventually runaway gas accretion settles in when the mass of the atmosphere becomes comparable to the core mass. This explains the flat behavior of the elapsed time in the later stages of the evolution, since this is when the atmosphere growth time continuously decreases, as seen in Figure \ref{fig:growthtime}. We note that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$.  %and the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude.

The grain opacity is one source of uncertainty in our model, as it may be be lower or higher than the standard dust opacity. Dust settling or grain growth result in a lower opacity, while a high metallicity of the host star yields an enhanced opacity. Lower opacity is possible to be achieved in our regime, since we consider low planetesimal accretion rates and therefore an atmosphere that is mostly depleted of planetesimals.  A decrease in opacity reduces the number of steps that photons need to diffuse (since it increases their mean free path), resulting in a higher luminosity. In Figure \ref{fig:LtvsMopacity}, we explore opacity laws in which the coefficient $\kappa_0$ from equation (\ref{eq:opacitylaw}) is reduced by factors of 10 and 100. All things equal, a lower opacity increases the luminosity and hence speeds atmosphere growth by the same factor as the opacity reduction (see equations \ref{eq:structd} and \ref{eq:delrad}).

 \begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/opacity_effect.pdf}
%\vspace{-0.5in}
\caption{Luminosity and time evolution as a function of atmosphere mass for planets of fixed core mass $M_{\rm c}=5 M_{\oplus}$ and located at $a=10$ AU, for different opacities. The standard dust opacity is reduced by factors of 10 and 100, respectively. A lower opacity accelerates atmosphere growth.}
\label{fig:LtvsMopacity}
\end{figure}

%\subsection{Growth Time \textbf{temporary title}}
%\label{growthtime}

@@

We have shown in section \ref{profiles} that growth is continuously accelerated once the self-gravity of the atmosphere becomes important. This is due to both the decrease in luminosity and to the fact that less and less binding energy per unit of added mass is released as the envelope grows in mass, i.e. $-dE/dM$ becomes small. The right-hand side of equation (\ref{eq:coolingglobal}) therefore decreases with increasing atmosphere mass; in our calculations, we have found that it eventually becomes negative around the time when $M_{\rm{atm}} \sim M_{\rm{c}}$. This translates into a negative luminosity from the radiative-convective boundary. Physically, the luminosity $L$ represents the total heat flux through the surface defined by the radiative-convective boundary. The heat flux due to convective motions is always directed outwards and hence positive. The temperature gradient $dT/dr$ is negative, as the temperature decreases towards the outside, and the heat flux due to radiative diffusion is also positive as a result (see equation \ref{eq:structd}). This implies that, in the absence of planetesimal accretion and internal heat sources, there is no physical mechanism that can justify a negative luminosity. The quasi-static, constant luminosity approximation breaks down at this point and a more detailed, time-dependent model is necessary. However, we will show in section \ref{critical} that the existence of this unphysical result does not affect our estimates of the atmosphere evolution time or critical core mass, as the unphysical regime occurs after the slowest growth stage. %However, we have shown in section \ref{growthtime} that in all cases presented here, growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and the estimate of the time to initiate runaway gas accretion is therefore insensitive to the exact choice of crossover atmosphere mass to order of magnitude.

@@

\section{Critical Core Mass}
\label{critical}

@@

Standard static calculations of giant planet formation that consider planetesimal accretion as the main source of luminosity (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{rafikov06}) assume that the atmosphere is at all times in a steady state at which the luminosity due to planetesimal accretion is fully radiated away by the atmosphere. The core and the atmosphere grow simultaneously, and the mass of the envelope is a function of the core mass. As the atmosphere mass becomes comparable to the mass of the solid core, hydrostatic balance no longer holds and runaway gas accretion commences. Therefore, for a given set of disk conditions, there is a maximum mass the core can have before the onset of unstable accretion: this is defined as the critical core mass. %a single value for the critical core mass $M_{\rm{crit}}$ is found, when the mass of the gaseous envelope is equal to the mass of the core. This is the maximum mass the core can have before unstable gas accretion commences.

In our model the atmosphere is no longer in steady state because energy balance requires gas accretion, and the atmosphere contracts on a Kelvin-Helmholtz time scale. We have shown in section \ref{profiles} that atmosphere growth is slowest before $M_{\rm{atm}} \sim M_{\rm c}$, and so the estimate of the time to initiate runaway gas accretion is insensitive to the exact choice of crossover atmosphere mass to order of magnitude. We therefore choose the effective crossover mass, past which hydrostatic balance no longer holds, to be the minimum between $M_{\rm{c}}$ and the $M_{\rm{atm}}<M_{\rm c}$ at which our model becomes unphysical according to equation (\ref{eq:dti}), noting that both of these occur after the slow accretion phase for all our parameter choices. We define the \textit{crossover time} $t_{\rm{co}}$ as the time elapsed until the atmosphere mass becomes equal to the effective crossover mass. At $t_{\rm{co}}$, runaway accretion commences.

In this section we explore the dependence of the crossover time on core, disk and gas parameters, and determine the minimum core mass needed to initiate runaway gas accretion within the lifetime of the protoplanetary disk. In section \ref{Mct} we show how the crossover time varies with core mass for fixed disk parameters, demonstrating that higher mass cores achieve runaway accretion more quickly. We explore how the crossover time is affected by by the mean molecular weight of the nebular gas in section \ref{muopacity} and by the disk temperature and pressure in section \ref{TPeffects}. Finally, we determine the minimum core mass necessary for a giant planet to form before the nebular gas dissipates in section \ref{critcore}; we define this minimum mass as the \textit{critical core mass} for runaway accretion to commence.

@@

\subsection{Crossover Time as a Function of Core Mass}
\label{Mct}

We begin by generating atmosphere evolution profiles for cores of different masses under a fixed set of disk conditions. Figure \ref{fig:tvsM} displays $t_{\rm{co}}$ for $M_{\rm c}$ between 5 and 14 $M_{\oplus}$ at $a=10$ AU in our fiducial disk. The envelope grows faster if the core is more massive, as it can bind a gaseous atmosphere more quickly due to larger gravity. This can also be seen from the analytical results in equation (\ref{eq:tcoolan}). As such, there is a minimum core mass required to form a planet within a fixed crossover time, e.g. the typical lifetime of the protoplanetary disk, which we defined above as the critical core mass. We defer numerical estimates of the critical core mass to section \ref{critcore}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.52\textwidth]{../../figs/ModelAtmospheres/RadSelfGravPoly/PaperFigs/cumul_coolingtime_vs_Matm_10au_mu235.pdf}
%\vspace{-0.5in}
\caption{Time to grow an atmosphere of mass $M_{\rm{atm}}$ for cores with fixed masses between $5 M_{\oplus}$ and $14 M_{\oplus}$ at $a=10$ AU in our fiducial disk. The circles mark the crossover time where $M_{\rm{atm}} \sim M_{\rm c}$. The numbers are labeling the core mass in Earth masses. A larger core mass results in a shorter crossover time.}
\label{fig:tvsM}
\end{figure}

\subsubsection{Role of Mean Molecular Weight}
\label{muopacity}

%Figure \ref{fig:tvsMcomp} reproduces the crossover time as a function of core mass at $a=10$ AU from Figure \ref{fig:tvsM}.

 In this work we build evolutionary atmosphere profiles assuming an ideal gas equation of state with constant mean molecular weight and adiabatic index. A realistic gas mixture, however, has variable adiabatic gradient and molecular weight, due to non-ideal effects and processes like dissociation or ionization. Future work will include realistic equation of state tables, and we expect quantitatively different results. In this study we analyze only the effect of the mean molecular weight of the nebular gas on the crossover time. In Figure \ref{fig:tvsM} we compare our standard model given by $\delad=2/7$ and $\mu=2.35$ from Figure \ref{fig:tvsM} with a model with the same adiabatic index but a different molecular mass, $\mu=2$ (i.e., purely molecular hydrogen). For comparison, we also plot the results predicted by the analytic, non self-gravitating model described in section \ref{analytic}. However, we saw in equation (\ref{eq:tcoolan}) that the crossover time as predicted by the analytic results is too long. Here we are investigating whether the analytic model predicts the correct scalings as opposed to the actual absolute results, when compared to the numerical model. We therefore estimate the analytic crossover time as the time needed for the atmosphere to reach an effective crossover mass $M_{\rm{atm}}=f M_{\rm c}$, where $f = 0.13$. The scaled analytic results are in good agreement with the numerical models. A lower mean molecular weight translates into a longer crossover time for the same core mass, since an atmosphere composed of heavier molecules has a smaller scale height, and therefore a weaker pressure gradient to oppose self-gravity.

\subsubsection{Disk Temperature and Pressure}
\label{TPeffects}

A planet's location in the protoplanetary disk affects its crossover time only because the disk temperature and pressure vary with stellocentric distance. Here, we study the separate effects of disk temperature and pressure on the time evolution of the atmosphere. At a fixed distance in the disk $a=10$ AU and for a core mass $M_{\mathrm{c}}=5 M_{\oplus}$, we first fix the disk pressure $P_{\rm{d}}$ at the value given by equation (\ref{eq:Pd}) and vary the disk temperature $T_{\rm{d}}$. We then fix the disk temperature at the value given by equation (\ref{eq:diskb}) and vary $P_{\rm{d}}$. The resulting variation in crossover time is plotted in Figure \ref{fig:TPeffects}. For comparison, we also plot the analytic results, using the same fractional mass scaling as in subsection \ref{muopacity}, $f=0.13$,
